---
title: "Homework Exercise"
---

<head>

```{=html}
<script src="https://kit.fontawesome.com/ece750edd7.js" crossorigin="anonymous"></script>
```

</head>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

::: objectives
<h2><i class="far fa-check-square"></i> Learning Objectives</h2>

-   Import and structure your data for plotting
-   Create an informative and visually appealing graphic
-   Use `tidyverse` packages for data analysis
:::

<br>

## Task

Write an R Script to import, restructure and plot a dataset of your choice. You can use the case study below, find a dataset online or even simulate your own dataset. The script should be well commented and include explanations of each step. You should also include a brief description of the dataset and the question you are trying to answer with your analysis.

1. Import the dataset using the appropriate function from the `readr` package.
2. Restructure the dataset using functions from the `tidyr` or `dplyr` packages to get it into a tidy format.
3. Create a plot using `ggplot2` to visualize the data and answer your research question.

## Case Study: COVID-19 Data Analysis

The COVID-19 pandemic has had a profound impact on global health and economies. Analysing data related to COVID-19 can provide insights into the spread of the virus, the effectiveness of interventions, and the impact on different populations. For this task, you will analyse a dataset containing COVID-19 case reports from around the world in 2020.

1. Download the [COVID-19 dataset](https://www.kaggle.com/datasets/imdevskp/corona-virus-report/code?datasetId=494766&sortBy=voteCount) from Kaggle
2. Import the table `covid_19_clean_complete.csv` into R

This table contains daily reports of COVID-19 cases, deaths, and recoveries for different countries and regions. The columns include:

- `Province/State`: The province or state of the report (if applicable)
- `Country/Region`: The country or region of the report
- `Lat`: The latitude of the location
- `Long`: The longitude of the location
- `Date`: The date of the report
- `Confirmed`: The total number of confirmed cases
- `Deaths`: The total number of deaths
- `Recovered`: The total number of recoveries
- `Active`: The total number of active cases
- `WHO Region`: The World Health Organization region of the report

3. Try to recreate the plot below using `ggplot2`:

```{r, fig.width=8, fig.height=6,echo=F}
library(tidyverse)
t <- read_csv("data/covid_19_clean_complete.csv",col_names = T)

latest_date <- max(t$Date)

tm <- t |> 
  group_by(`Country/Region`, Date, `WHO Region`) |> 
  summarise(Confirmed = sum(Confirmed),
            Deaths = sum(Deaths),
            Recovered = sum(Recovered),
            Active = sum(Active)) |> 
  rename(Country = `Country/Region`, WHO_Region = `WHO Region`) |># rename `Country/Region` to Country for easier typing
  ungroup()

countries <- c("Brazil", "Italy", "China", "New Zealand", "United Kingdom", "US")

library(ggrepel)
tm |> filter(Date == latest_date) |> 
  ggplot(aes(x = Confirmed, y = Deaths, colour = WHO_Region)) +
  geom_point() +
  geom_label_repel(
    data = tm |> filter(Country %in% countries, Date == latest_date),
    aes(label = Country), 
    size = 4,nudge_x = 3,
    show.legend = FALSE) + ## Add labels to the points
  scale_x_log10() + ## Use a log scale to better visualise the trends
  scale_y_log10() +
  scale_colour_brewer(palette = "Dark2") + ## Use a colour palette for better aesthetics
  labs(title = "Total Deaths vs Confirmed Cases",
       x = "Total Confirmed Cases (log scale)",
       y = "Total Deaths (log scale)",
       colour = "WHO Region") +
  theme_minimal()
```


::: hints
<h2><i class="far fa-flag"></i> Hints</h2>

-  You will need to combine data from different regions into one country using the `group_by()` and `summarise()` functions. Some countries have regions that are oversees territories. For simplicity, combine them all into one country.
-  Use the **ggrepel** package to add labels to your plot.
-  The colour palette used here is *Dark2* from the `RColorBrewer` package.
:::

