---
title: "Homework exercise"
---

<head>

```{=html}
<script src="https://kit.fontawesome.com/ece750edd7.js" crossorigin="anonymous"></script>
```

</head>

```{r global_options, include=FALSE}
knitr::opts_chunk$set(warning=FALSE, message=FALSE)
```

::: {.box .objectives}
<h3><i class="far fa-check-square"></i> Learning Objectives</h3>

-   Use publicly available RNA-seq data
-   Perform DE analysis with DESeq2
:::

## Task

There are many RNA-seq datasets available in the public domain. Find an RNA-seq dataset of interest to you and perform a DE analysis using DESeq2. You can use the code from the previous exercises as a template for your analysis.

You can search for datasets on [GEO](https://www.ncbi.nlm.nih.gov/geo/) or [SRA](https://www.ncbi.nlm.nih.gov/sra). You can also use datasets from the [Bioconductor ExperimentHub](https://bioconductor.org/packages/release/bioc/html/ExperimentHub.html).

The two datasets below are available as Bioconductor packages. Read the package vignettes for each dataset to learn more about the experimental design and how to access the data in R.

### Airway dataset

The [airway](https://www.bioconductor.org/packages/release/data/experiment/html/airway.html) dataset contains RNA-seq data from human airway smooth muscle cells treated with dexamethasone (a synthetic glucocorticoid) or left untreated.

The airway dataset is provided as a `summarizedExperiment` object, which can be imported into DESeq2 using the `DESeqDataSet()` function.

```{r,eval=F}
## Import the airway dataset into DESeq2
library(airway)
data(airway)

dds <- DESeqDataSet(airway, design = ~ cell + dex)
```

### Pasilla dataset

The [pasilla](https://bioconductor.org/packages/release/data/experiment/html/pasilla.html) dataset contains RNA-seq data from Drosophila melanogaster cells treated with dsRNA against the splicing factor pasilla or left untreated.

The pasilla dataset is provided as a set of count matrices and sample annotation files, which can be imported into DESeq2 using the `DESeqDataSetFromMatrix()` function.

```{r,eval=F}
## Import the pasilla dataset into DESeq2
library(pasilla)

## Count data file
pasCts <- system.file("extdata",
                      "pasilla_gene_counts.tsv",
                      package = "pasilla")

## Annotation file
pasAnno <- system.file("extdata",
                       "pasilla_sample_annotation.csv",
                       package = "pasilla")

## Read in the count data and annotation
countData <- as.matrix(
  read.csv(pasCts,
           sep = "\t",
           row.names = "gene_id")
)

colData <- read.csv(
  pasAnno,
  row.names = 1
)

colnames(countData)
colData
```

We need to fix the row names of the count data to match the sample names in the annotation file.

```{r,eval=F}
library(tidyverse)
## Remove "fb" from the row names of the count data
rownames(colData) <- rownames(colData) |> str_remove("fb")

## Sort the annotation file by condition so that the untreated samples are first
colData <- colData |> arrange(desc(condition))

dds <- DESeqDataSetFromMatrix(
  countData = countData,
  colData = colData,
  design = ~ condition
)

## Set untreated as the reference level for the condition factor
dds$condition <- relevel(dds$condition, ref = "untreated")

```

## Output

-   Quality control plots to assess quality and reproducibility of the data
-   A table of differentially expressed genes with log2 fold changes and adjusted p-values
-   A volcano plot to visualize the results of the DE analysis
-   A brief interpretation of the results, including any interesting genes or pathways that were identified

## Further resources

Use these tutorials for further guidance:

-   [DESeq2 tutorial](http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)
-   [HBC Differential expression workshop](https://hbctraining.github.io/Intro-to-DGE/schedule/links-to-lessons.html)

